---
title: "Exploring San Francisco Bay Area's Bike Share System"
author: "Tony Lin"
date: "August 29, 2018"
output: html_document
---

Congested streets and slow-crawling traffic are a fact of life in many metropolitan areas, such as New York City, Los Angeles, and Chicago. Bike sharing is an innovative solution to combat such problems, and it works by dispersing a large fleet of publicly-available bikes throughout crowded cities for personal transport. Implemented in 2013, Ford GoBike is the first bike sharing system introduced in the US West Coast. Its 540 stations and 7,000 bikes sprawl across five cities in the San Francisco/Bay Area. As an avid GoBike user and a data fanatic, I am excited to explore the trip data that has been collected in 2018 using R.

```{r echo=FALSE}
setwd("C:/Users/Tony Lin/Desktop/DataScience-tutorial/fordgobike/")
```


## Data Acquisition

First, I downloaded the [system data](https://www.fordgobike.com/system-data) from the Ford GoBike website and saved it in my working directory. Currently, eight data files are available online, a collective one for 2017 and one for each month of 2018 up to July. In this analysis, I will solely focus on data that has been generated in 2018.

```{r}
# List of data files
file_names = list.files(pattern = ".csv")
file_names
```

To read the files into R, I iterate through `file_names` using the `lapply` function:

```{r}
# Read data (this will take a second)
df_list = lapply(file_names,
                 function(x) read.csv(x, stringsAsFactors = FALSE))

# Assign name to data frames
names(df_list) = file_names
```


## Data Organization and Cleaning

### Aggregate Data into a Single Data Frame

Let's look at the headers of the first three data frames in `df_list`:

```{r}
# Extract column names from the first 3 data frames
head(sapply(df_list, names), 3)
```

Information on ride duration, starting and ending time/location, and usership are provided. Notice that the data contains a column called `"bike_share_for_all_trip"`, which tracks members who are enrolled in the Bike Share for All program for low-income residents.

We will collapse the data frames into one and examine the data structure:

```{r warning=FALSE, message=FALSE}
# Cast certain columns as numeric to enable row binding
df_list[[6]]$start_station_id = as.numeric(df_list[[6]]$start_station_id)
df_list[[6]]$end_station_id = as.numeric(df_list[[6]]$end_station_id)
df_list[[7]]$start_station_id = as.numeric(df_list[[7]]$start_station_id)
df_list[[7]]$end_station_id = as.numeric(df_list[[7]]$end_station_id)

# Bind data frames by rows
library(dplyr)
df = bind_rows(df_list)

# Glimpse at df
glimpse(df)
```

### Format Data Types

Data cleaning is essential for downstream analysis. This entails formating the columns into appropriate data types and extracting essential data. For instance, I will isolate the month from `start_time` for visualizing bike usage over time. In addition, I will convert `member_birth_year` into `member_age`. FInally, the `character` columns should take on type `factor`.

```{r warning=FALSE}
# Extract month from start_time
df = df %>%
  mutate(start_month = sub(" .*", "", start_time)) %>%
  mutate(start_month = as.POSIXct(start_month)) %>%
  mutate(start_month = format(start_month, "%B"))

# Extract hour from start_time
library(stringr)
df = df %>%
  mutate(start_hour = sub("^.{11}", "", start_time)) %>%
  mutate(start_hour = str_extract(start_hour, "^[0-9]{2}"))

# Convert birth year to age
df = df %>%
  mutate(member_age = 2018 - member_birth_year)

# Convert characters to factors
character_vars = which(sapply(df, is.character))
df[character_vars] = lapply(df[character_vars], as.factor)

# Remove columns not used in the analysis
df = select(df, -c(start_time, end_time, start_station_id, end_station_id, bike_id, member_birth_year))

# Glimpse at cleaned df
glimpse(df)
```

The cleaned data frame contains 12 descriptors on over a million rides.


## Data Analysis

### Bike Usage by Month in 2018

Let's examine the GoBike usage over time in 2018:

```{r}
# Organize month
df$start_month = factor(df$start_month, 
                        levels = c("January", "February", "March", "April", "May", "June", "July"))

# Count usage by month
month_counts = table(df$start_month)

# Plot barplot
bar = barplot(month_counts,
              ylim = c(0, 220000),
              xlab = "Months in 2018",
              ylab = "Number of Rides",
              main = "GoBike Usage by Month")
text(x = bar, y = month_counts,   # Add labels
     label = month_counts, pos = 3, cex = 0.8, col = "red")  
```

The plot shows a consistent growth in the number of rides over the months. Six months into 2018, the bike usage more than doubled. In addition, there's a large increase in the number of rides from April to May, which may be associated with warmer weather in the summer months.

### Bike Usage by Customers Versus Subscribers

Next, I will stratify the barplot by `user_type` to reveal information on the typical user.

```{r}
# Plot barplot
library(ggplot2)
ggplot(df, aes(start_month)) + 
  geom_bar() +
  facet_grid(. ~ user_type) +
  xlab("Months in 2018") + 
  ylab("Number of Rides") +
  ggtitle("Comparison of Usage Between Customers and Subscribers") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))   # Rotate x-axis labels
```

It is evident that the majority of the rides are made by subscribers of the monthly or yearly plan while only a fraction (~`r round(sum(df$user_type == "Customer") / nrow(df) * 100)`%) come from single-use customers. Proportionally, the number of users in both categories spiked from April to May, which reflects our initial finding.

### Bike Usage by Gender

In my experience, the bikes are predominately operated by males. Let's test that intuition and see whether the gender composition of users evolves over time:

```{r}
# Organize data by gender
df_gender = df %>%
  filter(member_gender %in% c("Male", "Female")) %>%   # Isolate "Male" and "Female" genders
  mutate(member_gender = droplevels(member_gender))   # Drop unused levels

# Plot barplot
ggplot(df_gender, aes(start_month, fill = member_gender)) +
  geom_bar(position = "fill") +
  xlab("Months in 2018") + 
  ylab("Proportion of Rides") +
  ggtitle("Bike Usage by Gender over Time") +
  guides(fill = guide_legend(title = "Gender")) +
  theme_minimal()
```

There are roughly three times as many male as female users, and the proportions have not changed over time.

### Bike Usage by Hour

```{r}
# Organize usage by hour
hour = df %>%
  group_by(start_hour) %>%
  summarise(total = n())

# Plot barplot
ggplot(hour, aes(x = start_hour, y = total, group = 1)) + 
  geom_line() +
  geom_point(color = "red") +
  xlab("Time in Hours") + 
  ylab("Number of Rides") +
  ggtitle("Number of Rides by Hour") +
  theme_minimal()

```


### Distribution of the Age of GoBike Users




### Distribution of Ride Duration




### Mapping GoBike Stations

Most used stations?

Peak usage?


## Closing Remarks

[Citi Bike](https://www.citibikenyc.com/system-data) is the Ford GoBike for New York City. Its system is more established and its data date back to 2013, which means more data to play with. I encourage you to 

